Bipartite Fidelity and Loschmidt Echo of Bosonic Conformal Interface

### 2017-05-08 free_boson_lattice.tex
[draft] general formulation revision 1 free boson lattice -> appendix?

{\bf\color{red} Say motivation: in low energy go to CFT}

{\bf\color{red} Say goal: Demonstrate the S matrix is conformal invariance}

>>>>

In this section, we consider a lattice model with bosonic interface at the center\cite{peschel_exact_2012,calabrese_entanglement_2012}, which reduces to the one considered in Sec.~\ref{sec_sub:general_formulation} in the continuum limit\cite{sakai_entanglement_2008}. Therefore, it serves as a numerical tool to check our analytic results in Sec.~\ref{sec:analytic_numerics}.

% Despite its simpleness, such discrete bosonic defect can be considered in more general quadratic bosonic\cite{PhysRevA.74.022329}, or fermionic\cite{PhysRevB.64.064412} systems. It also sheds light on the possible application of our CFT results in condensed matter experiments\cite{Rohringer2015,Trotzky2012}.

\todo[inline]{I think the reference cited here are irrelevant.}

We consider two harmonic chains with bosonic field $\phi_i$ and conjugate momentum $\pi_i$ on lattice site $i$. The left and right chains are connected between site $0$ and $1$ with the Hamiltonian
\begin{eqnarray}\begin{aligned}
H = \frac{1}{2} \sum_i \pi_i^2  +  \frac{1}{2} \sum_{i\ne 0 }  ( \phi_i - \phi_{i+1} )^2  +  \frac{1}{2} \begin{pmatrix}  \phi_0, \phi_1 \end{pmatrix}
\begin{bmatrix}
1 + \Sigma_{11}  & \Sigma_{12} \\
\Sigma_{21} &  1 + \Sigma_{22} \\
\end{bmatrix}
\begin{pmatrix}
\phi_0 \\
\phi_1 
\end{pmatrix}
\end{aligned}\end{eqnarray}
where the matrix $\Sigma$ parameterize two-site interaction. In fact $\Sigma$ gives a simple relation with the scattering matrix defined in Sec.~\ref{sec_sub:general_formulation}, and thus the Hamiltonian is the lattice regularization of the bosonic interface discussed there.

\todo[inline]{Mao: I suggest to relocate this part to an appendix, and maybe more details can be filled in, especially scale invariance condition imposed on the $\Sigma$. In the main text, just cite the relation between $\Sigma$ and $S$ and proceed}

To find the scattering matrix, we set up the plane wave scattering problem across the interface using ansatz for two semi-infinite chain (the use of $(n-1)$ in $\phi_n^B$ simplifies the calculation)
\begin{equation}
\label{eq:ansatz}
\phi_n
= \left\lbrace
  \begin{aligned}
    \phi_n^A &= A_{-} e^{i \omega t  - inka}  + A_{+} e^{i \omega t  + inka}  & \quad  n \le 0 \\
    \phi_n^B &= B_{-} e^{i \omega t  - i(n-1)ka}  + B_{+} e^{i \omega t  + i(n-1)ka} & \quad n \ge 1 \\
  \end{aligned} \right. 
 \quad 
\end{equation}
where $a$ is the lattice constant. The dispersion of the system is $\omega = \left|2\sin\frac{k}{2}\right|$ which implies the system is gapless. The interface will not break the criticality of the harmonic chain\cite{peschel_exact_2012}. Upon plugging Eq.~\eqref{eq:ansatz}-\eqref{eq:dispersion} into the equation of motion, we have the following relation between the incoming and outgoing amplitudes
\begin{equation}
\label{eq:discrete_S}
\begin{pmatrix}
A_{-} \\
B_{+}\\
\end{pmatrix}
=-
\begin{bmatrix}
\Sigma_{11} + e^{-ika} & \Sigma_{12} \\
\Sigma_{21} & \Sigma_{22}  + e^{-ika}  \\
\end{bmatrix}^{-1}
\begin{bmatrix}
\Sigma_{11} +e^{ika} & \Sigma_{12}\\
\Sigma_{21} & \Sigma_{22} + e^{ika}
\end{bmatrix}
\begin{pmatrix}
A_{+}\\
B_{-}\\
\end{pmatrix}
\end{equation}
and the scattering matrix can be solved as
\begin{equation}
  S = \frac{-1}{ \det \Sigma  + e^{-ika} \text{tr} \Sigma   + e^{-2ika}}
\begin{bmatrix}
\det \Sigma+ \Sigma_{11} e^{-ika} + \Sigma_{22} e^{ika}+1  & -2i \sin ka \Sigma_{12}  \\
-2i \sin ka \Sigma_{21} &  \det \Sigma+ \Sigma_{11} e^{ika} + \Sigma_{22} e^{-ika}+1\\
\end{bmatrix}
\end{equation}
A necessary condition for the scattering matrix to be scale invariance is that the ratio $|S_{11}/S_{21}|$ is independent of $k$. This implies 
\begin{equation}
\label{eq:Sigma_condition}
\det \Sigma = -1, \, \, \text{tr} \Sigma = 0
\end{equation}
Therefore, we have the scale invariant S-matrix as
\begin{equation}
S = \frac{1}{1 - e^{-2ika } } ( -2i \sin k ) \Sigma
 = - e^{ika} \Sigma
\end{equation}
In the continuum limit where $a\rightarrow0$, the matrix $\Sigma$ can be parameterized as
\begin{equation}
\Sigma = -\lim_{a \rightarrow 0 } S = 
\begin{bmatrix}
\frac{\lambda^2- 1}{1 + \lambda^2} & \frac{-2\lambda }{1 + \lambda^2} \\
\frac{-2\lambda }{1 + \lambda^2} & \frac{1- \lambda^2}{1 + \lambda^2} \\
\end{bmatrix}
\end{equation}
in accordance to Eq.~\eqref{eq:Sigma_condition}.

{\bf\color{red}
Say in the continuum limit, what happens

say in the long time, what happens?

say the special case in the boundary
}

This is the matrix we will use for the numerics. The results will be presented in Sec.~\ref{sec:analytic_numerics}.

%%% Local Variables:
%%% TeX-master: "bCFT_paper"
%%% TeX-PDF-mode: t
%%% End:

### 2017-05-08 general_formulation.tex
[draft] general formulation revision 1 free boson lattice -> appendix?

We consider a conformal interface connecting two potentially different CFTs in $1+1$ dimensions. The interface is located at $x=0$ and characterized by the ``gluing condition''
\begin{eqnarray}\begin{aligned}
\label{eq:def_M}
\begin{pmatrix}
\partial_x\phi\\
\partial_t\phi
\end{pmatrix}_{x=0^-}
=M\begin{pmatrix}
\partial_x\phi\\
\partial_t\phi
\end{pmatrix}_{x=0^+}
\end{aligned}\end{eqnarray}
The interface is assumed to be penetrable, and the stress tensor is continuous across the defect. From the explicit expression of the stress tensor $T^{xt}=-\partial_x\phi\partial_t\phi$, one can show that\cite{bachas_permeable_2002} $M$ is an element of the Lorentz group $O(1,1)$. In terms of the boost parametrization, we have
\begin{eqnarray}\begin{aligned}
M_1(\theta)=\pm
\begin{bmatrix}
\lambda^{-1} & 0 \\
0 & \lambda
\end{bmatrix}\quad
M_2(\theta)=\pm
\begin{bmatrix}
0 & \lambda  \\
\lambda^{-1} & 0 
\end{bmatrix}
\end{aligned}\end{eqnarray}
where $\lambda=\tan\theta$ for $\theta\in\left[-\frac{\pi}{2},\frac{\pi}{2}\right]$.

We note the following special cases. For $\theta=0,\pm\pi/2$, $\lambda$ (or $\lambda^{-1}$) appears to be singular and the fields on either side of the defect cannot transmit to the other side. For example, $M_1(0)$ implies $\partial_x\phi(0^+)=\partial_t\phi(0^-)=0$ which correspond to Dirichlet boundary condition on the left, and Neumann condition on the right. Hereafter we shall denote it as `DN'. Similarly one can check that $M_1(\pm\pi/2),M_2(0),M_2(\pm\pi/2)$ correspond to ND, NN, DD respectively. For $\theta=\pm\pi/4$, on the other hand, $M(\theta)$ characterizes a perfectly transmitting defect. For example, there is effectively no defect in the case $M(\pi/4)$. For the other three cases, although the field $\phi$ may pick up a phase across the defect, the two counter propagating waves are still fully transmitted.
\todo[inline]{shall we follow permeable wall paper to denote them as PP, etc? it is a bit confusing though..}

It proves useful to rewrite Eq.~\eqref{eq:def_M} in the ``light cone'' coordinate $x^\pm\equiv t\pm x$, where the physical meaning of $\theta$ is most transparent. For clarity, we denote the field on the negative (positive) real axis as $\phi^1$ ($\phi^2$).
\todo[inline]{Tianci: Add in the right panel in Fig.~\ref{fig:cut-and-join} the field content?}
After some algebra, we have
\begin{eqnarray}\begin{aligned}
\label{eq:def_S}
\begin{pmatrix}
\partial_+\phi^1\\
\partial_-\phi^2
\end{pmatrix}
=S
\begin{pmatrix}
\partial_-\phi^1\\
\partial_+\phi^2
\end{pmatrix}
\end{aligned}\end{eqnarray}
where 
\begin{eqnarray}\begin{aligned}
\label{eq:S1_S2}
S_1(\theta)=\begin{bmatrix}
\cos 2\theta & \sin 2\theta \\
\sin 2\theta & -\cos 2\theta
\end{bmatrix}\quad
S_2(\theta)=\begin{bmatrix}
-\cos 2\theta & \sin 2\theta \\
-\sin 2\theta & -\cos 2\theta
\end{bmatrix}
\end{aligned}\end{eqnarray}
The scattering matrix $S_1$ and $S_2$ relate the incoming and outgoing waves at the defect. For example, $S_1(0)$ implies that $\partial_\pm\phi^{1}=\pm\partial_\mp\phi^{2}$ as discussed above. For other nontrivial values of $\theta$, the defect becomes partially-transmitting, and the transmission and reflection coefficients can be read off from the scattering matrices. In particular, these coefficients depend only on the parameter $\theta$ of the defect, not on the wavelength of the incoming wave, which is in contrast to the case in mesoscopic system.\cite{datta_1995} This is expected due to the scale invariance.

The problem of conformal interface can be studied in the framework of boundary CFT\cite{cardy_boundary_2004,cardy_conformal_1984} after we fold the field $\phi^2$ to the left half plane on top of $\phi^1$\cite{oshikawa_boundary_2010}. The boundary at $x=0$ becomes impenetrable for the folded system, and the resulting tensor theory admits a conformal boundary states for the bosons. The folding sends $\phi^2(x)\rightarrow\phi^2(-x)$ for $x<0$, therefore the gluing condition reads
\begin{eqnarray}\begin{aligned}
\partial_t(\cos\theta\phi^1-\sin\theta\phi^2)=0 \quad
\partial_x(\sin\theta\phi^1+\cos\theta\phi^2)=0 
\end{aligned}\end{eqnarray}
for the case $M=M_1(\theta)$. After performing explicit mode expansion for both $\phi^{1,2}$, we have
\begin{eqnarray}\begin{aligned}
\label{eq:rotation_a_basis}
\cos\theta a_n^1-\sin\theta a_n^2 &= +( \cos\theta\bar{a}_{-n}^1-\sin\theta \bar{a}_{-n}^2 ) \\
\sin\theta a_n^1+\cos\theta a_n^2 &= -( \sin\theta\bar{a}_{-n}^1+\cos\theta \bar{a}_{-n}^2 ) 
\end{aligned}\end{eqnarray}
where $n>0$ is an integer. $a_n^{1,2}$ ($\bar{a}_n^{1,2}$) annihilate the left (right)-moving modes of $\phi^{1,2}$, and $a_{-n}^{1,2}$ ($\bar{a}_{-n}^{1,2}$) are the corresponding creation operators. Eq.~\eqref{eq:rotation_a_basis} is essentially a rotation of basis, and the conformal boundary state reads\cite{oshikawa_boundary_2010}
\begin{equation}
\label{eq:bd_state}
\begin{aligned}
| B \rangle 
& =  \exp\Big\{ \sum_{n > 0 } \frac{1}{n}
\begin{pmatrix}
a_{-n}^1\\
a_{-n}^2\\                              
\end{pmatrix}
S_1
\begin{pmatrix}
\bar{a}_{-n}^1  \bar{a}_{-n}^2
\end{pmatrix} \Big\} |0\rangle
\end{aligned}
\end{equation}
where $S^1$ is precisely the scattering matrix in Eq.~\eqref{eq:def_S}. As one can explicitly check, Eq.~\eqref{eq:bd_state} satisfies the condition for the conformal boundary state\cite{oshikawa_boundary_2010}. The calculation for the case $M=M_2(\theta)$ is completely analogue. Eq.~\eqref{eq:bd_state} will be used extensively in the boundary state calculation of fidelity and Loschmidt echo in Sec.~\ref{sec:analytic_numerics}.

>>>>

The general constraint we can impose on an interface is that the momentum flow (or current flow if exists) should be continuous. An equivalent way of viewing it (and will be used later) is to fold one of the theories on top of the other. The interface now located on the boundary of the tensor theory becomes impenetrable, and momentum flow ends here. This folding picture leads to the concept of boundary states in the framework of boundary CFT\cite{cardy_boundary_2004,cardy_conformal_1984}. In the Boson theory we are interested in, the tensor theory has $c = 2$ and the general classification the boundary state is still an open questions\cite{affleck_quantum_2001}.

There are however many successful attempts to construct a subset of those bosonic boundary states. One way to do is to impose current operator rather than Virasolo generator constraints, which dates back to the idea of the Ishibashi state\cite{ishibashi_boundary_1989} and can be applied to the multi-component boson with a general compactification lattice\cite{affleck_quantum_2001,oshikawa_boundary_2010,quella_reflection_2007}. Other example utilizes the fusion algebra to generate new boundary states from the known ones\cite{affleck_quantum_2001,bachas_fusion_2008}.

We here follow the presentation in Ref.~\onlinecite{bachas_permeable_2002} that begins with a classical boundary condition and then quantize it give the boundary state. The interface obtained is actually the same as using the current algebra constraints\cite{affleck_quantum_2001,oshikawa_boundary_2010,quella_reflection_2007}, but we feel that scattering process derived here and its relation to the discrete lattice model in Sec.~\ref{sec_sub:free_boson_lattice} are more intuitive.

Assume two free Bosons fields $\phi^1$ and $\phi^2$ live on the left and right half planes respectively. Their interface located at $x = 0$ are characterized by the "gluing condition"

\todo[inline]{Mao: I put $\partial_t$ on the top in my notes. The 'DN' 'ND' convention seems to be wrong here. I'll think about it.}
\begin{eqnarray}\begin{aligned}
\label{eq:def_M}
\begin{pmatrix}
\partial_t\phi^1\\
\partial_x\phi^1
\end{pmatrix}
=M\begin{pmatrix}
\partial_t\phi^2\\
\partial_x\phi^2
\end{pmatrix}
\end{aligned}\end{eqnarray}
where derivative should be understood in the appropriate left and right limits, for example $\partial_x \phi^1$ is evaluated at $ x = 0^-$. As argued, the momentum component of stress tensor is continuous across the interface. As a consequence $M$ is an element of the Lorentz group $O(1,1)$ and can be parameterized as
\begin{eqnarray}\begin{aligned}
M_1(\theta)=\pm
\begin{bmatrix}
\lambda^{-1} & 0 \\
0 & \lambda
\end{bmatrix}\quad
M_2(\theta)=\pm
\begin{bmatrix}
0 & \lambda  \\
\lambda^{-1} & 0 
\end{bmatrix}
\end{aligned}\end{eqnarray}
where $\lambda=\tan\theta$ for $\theta\in\left[-\frac{\pi}{2},\frac{\pi}{2}\right]$.

Several special choice of $\theta$ need to be noted. 
\begin{enumerate}
\item $\theta=0,\pm \frac{\pi}{2}$. In this case, $\lambda$ (or $\lambda^{-1}$) appears to be singular and the fields on either side of the defect cannot transmit to the other side, so the interface reduces to individual boundary conditions for the boson on the left and right half planes: they are combination of Dirichlet and Neumann boundary conditions. For example, $\lambda = 0$ for $M_1$ implies $\partial_x\phi^1 = \partial_t\phi^2 =0$, which means that Dirichlet boundary condition are imposed on the right and Neumann boundary condition on the left. Hereafter we shall denote this combination as 'DN'. Similarly $M_1(\pm\frac{\pi}{2}),M_2(0),M_2(\pm \frac{\pi}{2})$ correspond to DN, DD, NN respectively. 
\item $\theta = \pm \frac{\pi}{4}$. In this case, $M(\theta)$ characterizes a perfectly transmitting defect. For example, there is effectively no defect in the case $M( \frac{\pi}{4})$. For the other three cases, despite picking up a phase, the two counter propagating modes are still fully transmitted cross the interface. 
\end{enumerate}

The physical significance of $\theta$ will be manifest in the scattering process described below. We rewrite the Eq.~\eqref{eq:def_M} in the coordinates $t\pm x$ and $\partial_{\pm} = \partial_{t} \pm \partial_x $ derivative to extract the left and right going modes. For example, $\partial_{-} \phi^2$ will be a function $t - x$ only and hence represent a right going mode on the right half plane. It is then one of the scattering products that leave the interface. On the other hand $\partial_{-} \phi^1$ and $\partial_{+} \phi^2$ are modes that approaching the interface from their respective domain. We therefore establish the scattering relation 
\begin{eqnarray}\begin{aligned}
\label{eq:def_S}
\begin{pmatrix}
\partial_+\phi^1\\
\partial_-\phi^2
\end{pmatrix}
=S
\begin{pmatrix}
\partial_-\phi^1\\
\partial_+\phi^2
\end{pmatrix}
\end{aligned}\end{eqnarray}

\todo[inline]{Tianci: Add in the right panel in Fig.~\ref{fig:cut-and-join} the field content?}
and solve the $S$ matrix for the two cases of $M_1$ and $M_2$
\begin{eqnarray}\begin{aligned}
\label{eq:S1_S2}
S_1(\theta)=\begin{bmatrix}
\cos 2\theta & \sin 2\theta \\
\sin 2\theta & -\cos 2\theta
\end{bmatrix}\quad
S_2(\theta)=\begin{bmatrix}
-\cos 2\theta & \sin 2\theta \\
-\sin 2\theta & -\cos 2\theta
\end{bmatrix}
\end{aligned}\end{eqnarray}
\todo[inline]{Mao: please do not interchangably use defect or interface. We use interface throughout the text.}
For generic values of $\theta$, the interface is partially-transmitting, whose transmission and reflection coefficients can be read off from the scattering matrices. The fact of their wavelength independence is an indication of the interface's scale invariance.

We now work in the folding picture as shown in Fig.{\color{red} TODO}. The boundary at $x=0$ becomes impenetrable for the folded system, and the resulting tensor theory admits a conformal boundary state. The folding sends $\phi^2(x)$ to $\phi^2(-x)$ and hence the gluing condition becomes
\begin{eqnarray}\begin{aligned}
\partial_t(\cos\theta\phi^1-\sin\theta\phi^2)=0 \quad
\partial_x(\sin\theta\phi^1+\cos\theta\phi^2)=0 
\end{aligned}\end{eqnarray}
for the case $M=M_1(\theta)$.

If we quantize the Boson theory on the interface line $x = 0$, these gluing condition becomes an operator identity for the Boson creation and annihilation of the modes. We shall interpret these identities to be valid only when acting on the boundary states. The mode expansion of free Boson on $x = 0$\cite{di_francesco_conformal_1997} is
\begin{equation}
\phi(z, \bar{z} ) = \phi_0 - \frac{i}{4\pi g } \pi_0 \ln z \bar{z}  + \frac{i}{\sqrt{4\pi g} } \sum_{n \ne 0 } \frac{1}{n} \left(a_n z^n - \bar{a}_{-n } \bar{z}^{-n}   \right)
\end{equation}
where take $T$ to be time period and the following choice of the holomorphic and anti-holomorphic coordinates
\begin{equation}
z= e^{ 2\pi ( i\tau - x ) / T } \quad \bar{z} = e^{ -2\pi ( i\tau + x ) / T }
\end{equation}
We end up with a set of operator identities for each mode
\begin{eqnarray}\begin{aligned}
\label{eq:rotation_a_basis}
\cos\theta a_n^1-\sin\theta a_n^2 &= +( \cos\theta\bar{a}_{-n}^1-\sin\theta \bar{a}_{-n}^2 ) \\
\sin\theta a_n^1+\cos\theta a_n^2 &= -( \sin\theta\bar{a}_{-n}^1+\cos\theta \bar{a}_{-n}^2 ) 
\end{aligned}\end{eqnarray}
which is valid on the following boundary state
\begin{equation}
\label{eq:bd_state}
\begin{aligned}
| B \rangle 
& =  \exp\Big\{ \sum_{n > 0 } \frac{1}{n}
\begin{pmatrix}
a_{-n}^1\\
a_{-n}^2\\                              
\end{pmatrix}
S_1( \theta )
\begin{pmatrix}
\bar{a}_{-n}^1  \bar{a}_{-n}^2
\end{pmatrix} \Big\} |0\rangle
\end{aligned}
\end{equation}
where $S_1(\theta) $ is precisely the scattering matrix in Eq.~\eqref{eq:def_S}. The calculation for the case $M=M_2(\theta)$ is completely analogous. Eq.~\eqref{eq:bd_state} will be used extensively in the boundary state calculation of fidelity and Loschmidt echo in Sec.~\ref{sec:analytic_numerics}.
